---
title: "lab_04"
author: "Derek Willis"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# load your packages

library(tidyverse)
library(lubridate)
library(janitor)
library(refinr)


```

For this lab, we want to investigate spending by Maryland state and local candidates during the previous election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table & assign variable name
#clean column names
#date columns should be date datatypes

maryland_election_expenses <- read_csv("data/maryland_expenses.csv") |> 
#glimpse(maryland_election_expenses)
mutate(`Expenditure Date` = mdy(`Expenditure Date`)) |> 
clean_names()


maryland_election_expenses
```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

1. Create a dataframe of Media expenses where the name of the payee is not NA.

2. Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.

  
  **A1-A2**
  
  **I used Chat GPT to help with the function "write_csv."** This is the prompt I used: 
  "how do i fix this code to use the function "write_csv"? media_expenses <- maryland_election_expenses |> 
  mutate(expense_upper = str_to_upper(expense_category), payee_upper = str_to_upper(payee_name)) |> 
select("expense_upper", "payee_upper") |> 
  filter(expense_upper =='MEDIA' & payee_upper != "NA") |> 
 media_expenses <- write_csv("data/media_expenses.csv")" 
 

```{r}
#Create a dataframe of Media expenses where the name of the payee is not NA.

media_expenses <- maryland_election_expenses |> 
  mutate(expense_upper = str_to_upper(expense_category), payee_upper = str_to_upper(payee_name)) |> 
#select("expense_upper", "payee_upper") |> 
  filter(expense_upper =='MEDIA' & payee_upper != "NA") |> 
write_csv("data/media_expenses.csv")

media_expenses
``` 
 
   3. Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.

   4. When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
   
    5. Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A3-A5**
I could not figure out/remember how to manually combine values, like facebook and meta, together in Open Reine. 

   5. Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

   
```{r}

media_expenses_clean <- read_csv("data/media-expenses-refine.csv") |> 
#  select(committee_name, amount, payee_clean) |> 
  mutate(payee_merge=key_collision_merge(payee_clean)) |> 
  filter(
    str_detect(payee_merge, 'FACEBOOK') |
   str_detect(payee_merge, "META")) |> 
  group_by(committee_name) |> 
summarise(total=sum(amount)) |> 
arrange(desc(total))

media_expenses_clean

```
   

**Q2.** Let's do some research on the top committees from question 1. 

*Look up the top 10 spenders:*

Jain Ashwani Friends of
Franchot Anderson-Walker for Maryland Slate
Franchot Peter Friends Of	
Casa in Action PAC	
Boniface Billy Friends Of	
Baron Jon (Jonathan) for Maryland	
Jasani Salimah Friends of	
WFP National Pac - Non-Contribution Account	
King and Siri for Maryland Slate
Shuman Eve T. Friends of	

*If they are candidate committees, what offices were they running for?*

- Jain Ashwani Friends of | Ashwani ran for governor in 2022
- Franchot Anderson-Walker for Maryland Slate & 
  Franchot Peter Friends Of	| Anderson-Walker joined Peter V.R. Franchot's gubernational ticket as the candidate for        lieutenant governor. 
- Boniface Billy Friends Of	| Republican candidate for Harford County Executive 
- Baron Jon (Jonathan) for Maryland | ran for election for Governor of Maryland (D)
- Jasani Salimah Friends of	| Jasani ran for Balimtore City Board of Education 
- King and Siri for Maryland Slate | King and Siri were on the ticket together for Governor of Maryland. 
- Shuman Eve T. Friends of | Shuman ran for Prince George's County District 3 Councilmember


*If they were not candidates, what issues do they seem to care about?*

- Casa in Action PAC | an individual membership org mobilizing Black and brown voters to elect progressive leaders. Endorses canidates that advance priorities of Latinos and voters of color in elections. 
- WFP National Pac - Non-Contribution Account	| The Working Families Party is a multiracial party fighting for workers and people, and is challenging the two party system. 


Write some sentences answering those questions, then go the Facebook ad library and look for ads from the top spender. *What do you notice about the the information provided about those ads?*

All ads for Jan for Gov, paid by Friends of Ashwani Jain, were removed: "This content was removed because the disclaimer didn't followour policy or ads about social issues, elections or politics."

([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\%5Bdirection\%5D=desc&sort_data\%5Bmode\%5D=relevancy_monthly_grouped&media_type=all){.uri}) 

**A2.** 

The committee, "Jain Ashwani Friends of" supported Ashwani's run for governor. "Franchot Anderson-Walker for Maryland Slate" supported Franchot's gubernational ticket as the candidate or lieutenant governor, with Anderson-Walker. "Boniface Billy Friends Of" supported Boniace's run for Harford County Executive. "Baron Jon (Jonathan) for Maryland" supported Baron's run for Governor of Maryland. "Jasani Salimah Friends of" supported Jasani's run for Baltimore City Board of Education. "King and Siri for Maryland Slate" supported King and Siri as the ticket or Governor of Maryland. "Shuman Eve T. Friends of" supported Shuman's run for Prince George's County District 3 Councilmember

Casa in Action PAC is an individual membership organization mobilizing Black and brown voters to elect progressive leaders. The PAC endorses canidatesadvancing priorities of Latinos and voters of color in elections. The WFP National Pac - Non-Contribution Account supports the Working Families Party, a multiracial party challening the two party system while fighting for workers and people. 


**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>):

*who is missing from that Q1 list?*

Dan Cox and Gordana Schifanell, Robin Ciker and LeRoy Yegge, Joe Werner and Minh Thanh Luong, are all Republican tickets not included in the Q1 list. 

Rushern L. Baker III and Nancy Navarro, Douglas Gansler and Candace Hollingsworth, Ralph Jaffe and Mark Greben, Wes Moore and Aruna Miller, Tom Perez and Shannon Sneed, and Jeroma Segal and Justinian Dispenza. 

*Did those campaigns run Facebook ads?* 

*Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?*


**A3.** 

*who is missing from that Q1 list?*

Dan Cox and Gordana Schifanell, Robin Ciker and LeRoy Yegge, Joe Werner and Minh Thanh Luong, are all Republican tickets not included in the Q1 list. 

Rushern L. Baker III and Nancy Navarro, Douglas Gansler and Candace Hollingsworth, Ralph Jaffe and Mark Greben, Wes Moore and Aruna Miller, Tom Perez and Shannon Sneed, and Jeroma Segal and Justinian Dispenza. 

*Did those campaigns run Facebook ads?* 

- Dan Cox and Gordana Schifanell | YES
- Robin Ciker and LeRoy Yegge | NO
- Joe Werner and Minh Thanh Luong | NO
- Rushern L. Baker III and Nancy Navarro | YES
- Douglas Gansler and Candace Hollingsworth | YES
- Ralph Jaffe and Mark Greben | NO
- Wes Moore and Aruna Miller | YES
- Tom Perez and Shannon Sneed | YES
- Jeroma Segal and Justinian Dispenza | NO


*Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?*

Moore Wes For Maryland
Moore Miller For Maryland Slate

```{r}
media_expenses |> 
  select(committee_name, payee_name, amount) |> 
  mutate(type = case_when(
    str_detect(committee_name, "Moore  Wes For Maryland"),
    str_detect(committee_name, "Moore Miller For Maryland Slate"))) |> 
      group_by(committee_name)
    


```


```{r}

media_expenses |> 
  select(committee_name, payee_name, amount)  |>
  filter(
    committee_name == 'Moore Wes For Maryland' |
    committee_name == 'Moore Miller For Maryland Slate'
  ) #|> 
 # mutate(Moore_Miller = TRUE)
  
 # filter(committee_name == 'Moore Wes For Maryland' | committee_name == 'Moore Miller For Maryland Slate' )

```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** 

```{r}
```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5**
