---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? **A1.**

26 departments have offered the course with climate in the title, with at least one seat. 

The Atmospheric and Oceanic Science department offers the most. 
  


```{r}
umd_courses |>
  filter(str_detect(title, "Climate") & seats>0) |>
  select(department, title, seats) |> 
  count(department) |> 
  arrange(desc(n)) 

```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why? 

MD Institute for Applied Environmental Health. I know the department is doing really interesting things at UMD. I'm sure I'd learn a lot on the intersection of economics, public health, climate and policy. 


**A2.**

**Q3** Which departments offered the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?

**A3.**

a. Which departments offered the most courses in the Fall 2023 term?

  umd_courses |> filter(term == "202308" | term== "202309" | term == "202310" | term == "202311" | term == "202312") |> 
  select(department, syllabus_count, title) |> 
 group_by(department) |> 
  summarise(title_count = n()) |> 
  arrange(desc(title_count))
  

*Business and Management offered the most courses.* 


b. And which departments in that term have at least one syllabus posted?

  umd_courses |> filter(term == "202308" | term== "202309" | term == "202310" | term == "202311" | term == "202312") |> 
  filter(syllabus_count>0) |> 
  select(department, syllabus_count, title) |> 
 group_by(department) |> 
  summarise(total_syllabus_count = sum(syllabus_count)) |> 
  arrange(total_syllabus_count) |> 
distinct(department) |> 
  count()


*It would appear that 125 departments have at least one syllabus posted.*

*question* I am wondering whether I am misunderstanding what to include for the fall term?


```{r}
  
  umd_courses |> filter(term == "202308" | term== "202309" | term == "202310" | term == "202311" | term == "202312") |> 
  filter(syllabus_count>0) |> 
  select(department, syllabus_count, title) |> 
 group_by(department) |> 
  summarise(total_syllabus_count = sum(syllabus_count)) |> 
  arrange(total_syllabus_count) |> 
distinct(department) |> 
  count()

 
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe. 

Then create a new column for the month based on expenditure_date. 

Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name). 

Write a sentence describing the pattern of spending over time. 

**A4.**

The committee started to spend more money beginning in Jan, 2022, and ramped up the summer before the 2022 election in June and July, with the most spent in October, which I suppose was more or less the "last call" before the election took place. 

**AC_notes**

maryland_expenses_dateformat <- maryland_expenses |>
  mutate(expenditure_date = mdy(expenditure_date)) |> 
  mutate(month= floor_date(expenditure_date, "month")) |> 
  mutate(
    committee_upper = str_to_upper(committee_name)) |> 
  select(expenditure_date, month, committee_name, amount) |> 
  mutate(
    committee_upper = str_to_upper(committee_name)) |> 
    filter(str_detect(committee_upper, "MOORE")) |>
    
    
  
maryland_expenses_dateformat <- maryland_expenses |>
  mutate(expenditure_date = mdy(expenditure_date)) |> 
  mutate(month= floor_date(expenditure_date, "month")) |>
  group_by(month) |> 
  summarise(total_amount = sum(amount))

  

```{r}

maryland_expenses_dateformat <- maryland_expenses |>
  mutate(expenditure_date = mdy(expenditure_date)) |> 
  mutate(month= floor_date(expenditure_date, "month")) |> 
  mutate(
    committee_upper = str_to_upper(committee_name)) |> 
  select(expenditure_date, month, committee_name, amount) |> 
  mutate(
    committee_upper = str_to_upper(committee_name)) |> 
    filter(str_detect(committee_upper, "MOORE")) |>
group_by(month) |> 
  summarise(total_amount = sum(amount)) |> 
arrange(month)

```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. 

For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". 

Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).


**A5**

Total amount in-state: 77,723,146
Total amount out-of-state: 49,459,974

Difference: 28,263,172 

Results: $ 28,263,172  more was spent in-state than out-of-state. 


```{r}

library(scales)

maryland_expenses_spending_location <-maryland_expenses |> 
  select(amount, address) |> 
  mutate(
    address_upper = str_to_upper (address)) |> 
  mutate(
    In_state = case_when(
        str_detect(address, " Maryland ") ~ "In-state",
    .default = NA
    )
  ) |>  
   mutate(
    Out_of_state = case_when(
      !str_detect(address, " Maryland ") ~ "Out-of-state",
      .default = NA
    )
  ) |> 
  
summarise(
    In_state_expenses = sum(amount[In_state == "In-state"], na.rm = TRUE),
    Out_of_state_expenses = sum(amount[Out_of_state == "Out-of-state"], na.rm = TRUE),
    Difference =  In_state_expenses - Out_of_state_expenses
    )


```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. 


Q: I'd like to group/summarize the expense purpose column by type, and then organize by descending order to see where most of the money is going toward. 

**A6.**

The majority of the funds went toward "transfer to MD Registered political committees," followed by TV, which I presume means advertisements on TV. Next, I would like to learn how to account for all the missing "NA" expenses, as there is a good amount of money considered "NA," and so we do not know the amount of money that is not accounted for. 


```{r}

maryland_expenses_spending_purpose <-maryland_expenses |> 
select(payee_type, expense_purpose, amount) |> 
mutate(
  expense_purpose_upper = str_to_upper (expense_purpose)
) |> 
  group_by(expense_purpose) |> 
    summarise(total_amount = sum(amount)) |> 
arrange(desc(total_amount))
```
