---
title: "lab_01"
author: "Derek Willis"
date: "2023-01-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you?
**A1**
Engineering, Cooperative Education. I find it interesting that Engineering has almost 2x more seats (on average) than the next highest average number of seats,"First-Year Innovation & Research Experience"

I also find it interesting that some of the average seats are significantly lower than the total seats, which could be due to some classes being virtual, and lectures vs. labs and higher level classes, which tend to be smaller. Example:biology has total 1,474	seats, but the mean is ~8 seats. 

	
```{r}
umd_courses |> 
  group_by(department) |> 
  summarise(
    total_seats = sum(seats),
    mean_seats = mean(seats)
  ) |> 
  arrange(desc(mean_seats))
          
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first. 
**A2**

4,975

```{r}
umd_courses |> 
  group_by(term) |> 
  summarise(
    count_classes = n()
  ) |> 
  arrange(desc(count_classes))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most?

**Updated Q3** Of all the departments, which instructors teaches the most in their (primary) department? 

**A3**

Keith Yearwood

**Disclaimer** Chat GPT helped me with the code, which is why it looks different than what we did in class. The funcition "arrange" was not working on my end in this code, and Chat GPT told me to load the "dplyr" library. I am not sure why the function was working previously though, and not in this code. I was also running into error messages when I copied the code Laura used in class, and asked Chat GPT what I was doing wrong, and it outputted the code below. I think I just had some spaces missing in the code for "Instructor: TBA" ?


```{r}
umd_courses |> 
  group_by(instructors, department) |> 
  summarise(
    count_classes = n()
  ) |> 
  arrange(desc(count_classes)) |> 
  filter(!is.na(instructors) & instructors != "Instructor: TBA")
         
```


**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4**

I suppose initially having the data sheet organized by the known instructors (not including NA or Instructor: TBA), the number of classes they teach and by department. I am not as confident as I would have been if I wrote down the name of the answer said in class. I am 80% confident I am right after utilizing Chat GPT (for the first time). I was also able to ask Chat GPT to elaborate on some of the functions so help me better understand R, and the code written above. 


**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? 

**A5**
I updated my question to look at the number of classes offered per year, starting in 2018, to note any trends during the pandemic as it pertains to classes offered. I am having a difficult time figuring out how to aggregate the years though, so 2018 only appears once in the row. 

**Disclaimer**
I used Chat GPT to help write the code below.


```{r}
umd_courses |> 
  mutate(year= as.integer(substr(term, 1,4))) |> 
  group_by(year, term) |> 
  summarise(
    count_classes = n(),
  ) |> 
  arrange((term)) |> 
  filter(term >= 201801)
         
```
