---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
# Load janitor

library(tidyverse)
library(refinr)
library(janitor)
library(lubridate)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
# Load in the Maryland grants and loan data


md_payments <- read_csv("/Users/AlishaCamacho/Documents/GitHub/camacho-alisha-jour-772/lab_03(1)/data/State_of_Maryland_Payments_Data__FY2008_to_FY2024_V2.csv") |> 
  clean_names() |> 
  mutate(
    agency_name = str_to_lower(agency_name),
    vendor_name = str_to_lower(vendor_name)) |> 
  mutate(date = dmy_hms(date),
         date = date(date)) |>
  mutate(year = year(date))

md_payments |>
  group_by(year) |>
  summarise(count = n())

```

```{r}
md_grant_loan <- read_csv("/Users/AlishaCamacho/Documents/GitHub/camacho-alisha-jour-772/lab_03(1)/data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022_V2.csv") |>
  clean_names() |>
  mutate(date = mdy_hms(date)) |> 
  mutate(
    grantor = str_to_lower(grantor),
    grantee = str_to_lower(grantee), 
    description = str_to_lower(description)
  )

glimpse(md_grant_loan)

```

## Answer questions

#AC_Q1: how to separate date and time data types?

#Q1:

You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system. 

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? 

#A1:

66

```{r}

pg_schools <- md_payments |> 
  #select(vendor_name, vendor_zip, amount) |> 
  filter(
    str_detect(vendor_name, "prince george|pg"),
    #str_detect(vendor_name, "pg"), 
    str_detect(vendor_name, "school")
  )

pg_schools <- md_payments |>
  filter(str_detect(vendor_name, "p"),
         str_detect(vendor_name, "g"),
         str_detect(vendor_name, "s")) |>
  #select(vendor_name) %>%
  filter(
    str_detect(vendor_name, "prince george|pg"),
    str_detect(vendor_name, "school|board of education")
  ) 
  


```


Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

#A1

2021; pg schools didn't appear to get any funding in 2016. 

```{r}

pg_schools |> 
  group_by(fiscal_year) |> 
  summarise(amount = sum(amount)) |> 
  arrange(desc(fiscal_year))

```

```{r}
#looking for 2016 payments

md_payments |> 
  select(fiscal_year, vendor_name, amount) |> 
  filter(
    str_detect(vendor_name, "prince george")|
    str_detect(vendor_name, "pg"), 
    str_detect(vendor_name, "school")
  ) |> 
  #filter(fiscal_year = 2016)
  mutate(before_after_2016 = case_when(
    #~ seperates if from them
    #if fiscal year greater than 2016, then (~) assign greater
    fiscal_year > 2016 ~ "greater",
    fiscal_year < 2016 ~ "lesser",
    #if fiscal year equals 2016
    fiscal_year == 2016 ~ "is 2016"
  )) |>
  rename(year = fiscal_year)
  arrange(fiscal_year)

pg_schools_2016

= #assigns
== #this is an operator
>= #this is an operator
<= #this is an operator

```



Q3.1 You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

#A3.1
md_payments_capital_projects

```{r}
#use payments data
#select vendor_name, agency_name (capital projects @ umd), amount

md_payments_capital_projects <- md_payments |> 
  #select(agency_name, vendor_name, amount) |> 
  filter(
    str_detect(agency_name, "capital projects"),
    str_detect(agency_name, "university of maryland") |
    str_detect(agency_name, "college park") |
    str_detect(agency_name, "umd")
  )

md_payments_capital_projects

```

#Q3.2
Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

#A3.2

I would use open refine (which we hadn't learned about at the time).

3 vendors: a k j inc, 	a morton thomas & associates inc, a r marani inc

```{r}
# arrange to see duplicates by vendor and zip code

md_payments_capital_projects |> 
  group_by(vendor_name) |>
  arrange(vendor_name)

md_payments_capital_projects

```

Q4.1: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description.

#A4.1:

md_grant_loan_stem_cell

```{r}

#select amount, description (stem cell research), fiscal year, category
#filter for grants

md_grant_loan_stem_cell <- md_grant_loan |> 
  select(description, amount, fiscal_year, category) |> 
  mutate(category = str_to_lower(category)) |> 
  filter(category == "grant") |> 
  filter(
    str_detect(description, "stem cell") |
    str_detect(description, "stem cell research")
  )

md_grant_loan_stem_cell 
```
#Q4.2
Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

#A4.2

The research got the greatest financial boost in its first year, which makes sense. the grant money dropped a bit in 2021, which also makes sense given the pandemic, and then jumps back up in 2022. 


```{r}
#I used chat gpt to figure out how to convert to $. You'lll have to scroll to the bottom of the chat to see the back and forth on this one: https://chat.openai.com/share/bb988120-cdfe-48e7-88be-bfe6c7781ac8

library(scales)

md_grant_loan_stem_cell |> 
  group_by(fiscal_year) |> 
  summarise(amount = sum(amount)) |> 
  mutate(amount_in_dollars = dollar(amount)) |> 
  arrange(fiscal_year)

```


Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5:

#number of grantees = 17, umd is the #1 grantee


```{r}
#select zip code 20742, grantees, amount

md_grant_loan |>
  select(grantee, zip_code, amount, category, description) |> 
  filter(zip_code == "20742") |> 
  group_by(grantee) |> 
  summarize(total = sum(amount)) |> 
  arrange(desc(total))

```


```{r}

#NOTES DURING 1:1

function(string, filter_string) {
  
  # calls the object
md_payments |> 
  #we are just doing this to make the output simpler 
  select(vendor_name) |>
  #looks for rows in each column and checks to see if the pattern you fed it is inside that string
  mutate(str_detect = case_when(
    #if it is in that string, it says TRUE
    grepl(filter_string, string) == TRUE ~ TRUE,
    #if it is not in that string, it says FALSE
    .default = FALSE
  )) %>%
  #then string detect filters for only rows that are true
  filter(str_detect == TRUE)
  
}

str_detect(vendor_name = "prince")

```



```{r}

#NOTES DURING 1:1


#use combine function to combine arguements into a list
numbers <- c("1", "2", "3", "4", "5")

#this is an if statements that tests the equation inside, if it is true, it will do the instructions inside the brackets

"6" %in% numbers

```


```{r}

#NOTES DURING 1:1

important_fiscal_years <- c(2016, 2017, 2018)

md_payments |>
  filter(fiscal_year %in% important_fiscal_years)# |>
  #filter(fiscal_year == 2016 |
  #         fiscal_year == 2017 |
  #         fiscal_year == 2018)

#list of important state
full_name_states <- c("florida", "maryland")#, ... etc)

#function that turns state names into abbreviations

abrev_name_states <- abrev_function()



st == "Fl"|st == "MD"

state_data |>
filter(st %in% abrev_name_states)
#filter(st == "FL" | st == "MD")


```

