---
title: "lab_05"
author: "derek willis"
date: "2023-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

- Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

library(tidyverse)
library(refinr)
library(janitor)
library(lubridate)
library(readxl)
library(dplyr)

```

## Get Our PDF

We'll be working with the [911 overdose calls from Baltimore County](https://drive.google.com/file/d/1qkYuojGF_6WKFr5aNQxmewDzcKyOiJFr/view?usp=share_link). You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to extract the tables within it, export that to a CSV file, load it into RStudio and ask some questions.

## Extract Data from PDF Using Tabula

Start Tabula, then go to http://127.0.0.1:8080/ in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds or longer.

This PDF has a single table spread over multiple pages to extract. We're going to make a single dataframe from this table, exporting it to a CSV file that you will load into R. In Tabula, highlight the table and click the "Preview & Export Extracted Data" button. You may want to play with including or excluding the column headers - YOU SHOULD HAVE FIVE COLUMNS OF DATA.

Save the CSV (it should be called `tabula-Baltimore County; Carey, Samantha log OD.csv` by default) to your lab_05/data folder.

From there, you will need to read in the data, and add or fix headers if necessary. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)).


```{r}

"OD_Log_Baltimore" <- read_csv("tabula-Baltimore County; Carey, Samantha log OD.csv", col_names = FALSE)

OD_Log_Baltimore

```


## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers should not contain spaces and they should have meaningful names, not "x1" or something similar. How you do that is up to you, but you can use select() with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns. Importantly, you'll need to ensure that any columns containing a date actually have a date datatype. Our friend `lubridate` can help with this.

```{r}

"OD_Log_Baltimore" <- read_csv("tabula-Baltimore County; Carey, Samantha log OD.csv", col_names = FALSE) |> 
clean_names() |> 
  rename(datez = x1, time = x2, case_number = x3, EVTYP = x4, location = x5) |> 
  mutate(datez = mdy(datez))

OD_Log_Baltimore |> 

OD_Log_Baltimore

```

## Answer questions

Q1. Write code to generate the number of calls that occurred on each date. Which date in 2022 had the most overdose calls, and how many? Look at the total number of rows in your result and explore the range of dates - based on your result, do you believe there are any days with no overdose calls at all? Explain why or why not.

#A1. 

July 14 and September 4th both had the most overdose calls in the dataset at 23 calls. It would also appear that calls were made 353/356 days of the year in 2022 and 2023. So not all the days, but close. 


```{r}
#Disclaimer, I used Chat GPT to help with this code. Please refer to this conversation: https://chat.openai.com/share/02928840-7d49-463a-b357-aac1fef92953. I am not quite sure why the original code I was using did not work... 



OD_Log_Baltimore_count <- OD_Log_Baltimore #%>% 
 # filter(year(datez) == 2022) |> 
 # group_by(datez) 
  
  #group_by(dates) |> 
  #summarise(n = n()) |>
 # count(dates = "2022") |> 
  #count(dates = "2023") |> 
  #arrange(desc(dates))

#str(OD_Log_Baltimore_count)

#OD_Log_Baltimore_count


```



```{r}

OD_Log_Baltimore |> 
group_by(datez) |> 
  summarise(n=n())

OD_Log_Baltimore

```

Q2. You want to understand if there's a pattern in the day of the week that overdose calls are made. Add a column to your dataframe that displays what day of the week each date represents. You should search for how to do that using lubridate.

Then write code to calculate the number of calls for each day of the week, and add a column to that result that calculates the percentage of all calls that occurred on each day of the week (so you want a dataframe with the day of the week, total number of calls and the percentage of calls on that day out of the total number of all calls). Describe your findings to me.

#A2. 

It would appear the most calls were made on Sundays and Saturdays, which makes sense to me. This could be a combination of more people ODing on the weekends, not during work hours, and having more eyes to call in ODs, since more people are not working and may notice more people ODing in their surroundings. 


#Diclaimer, I used Chat GPT to help with the code. Here is the link to the conversation: https://chat.openai.com/share/48920ade-c206-4c7f-b417-b23c0dfc7741


```{r}

OD_Log_Baltimore_DOW <- OD_Log_Baltimore %>%
  mutate(day_of_week = wday(datez, label = TRUE)) |> 
  group_by(day_of_week) |> 
  summarise(count = n()) |> 
  mutate(percentage = (count/sum(count)) * 100)

OD_Log_Baltimore 

```

Q3. Now let's look at locations. Which ones have the most calls? How would you describe them (feel free to search for more information on them)? Is there anything about the structure of the original data that might make you less confident in the counts by location or date?

#A3. 

LOCATIONS: 

4540 SILVER SPRING RD
- residential 
- currently pending for sale as of Oct/15/2023

PC 06; 115 SUSQUEHANNA AV W	
- Baltimore county police department p3

PC 02; 6424 WINDSOR MILL RD
- Baltimore county police department precinct 2

330 LEEANNE RD	
- residential 


It would appear that the calls and locations are being made at police precincts, which may not be the most representative of the actual day and location of the incidents, and may just be when they were formally reported/documented. 


```{r}

OD_Log_Baltimore |> 
  group_by(location) |> 
  summarise(count = n()) |> 
  arrange(desc(count))

```

Q4. What's the best story idea or question you've seen as a result of the work you've done in this lab?

#A4. 

1) I cannot figure out what EVTYP means, and I really want to know what it means/stands for. 

2) In regards to a story ... I'd be interested in following up on what was happening at 4540 SILVER SPRING RD	, and why the property went up for sale. Is this connected to the OD calls? It could be nothing, but I'd be curious to learn more about what was going on there.  



