---
title: "lab_07"
author: "derek willis"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
library(refinr)
library(lubridate)
library(janitor)
rm(list = ls())

```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}

census_api_key("a0c53f3750a5b78a93b8b40993886ff655a3d0fd", install = TRUE, overwrite =TRUE)
readRenviron("~/.Renviron")
```
```{r}
readRenviron("~/.Renviron")
```


## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2021 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2021 5-year ACS (which is called "acs5") and save them to a dataframe

#disclaimer, I used Chat GPT to figure out how to format the code to load the variables: https://chat.openai.com/share/8338b2b7-ddfa-46fd-88e4-00d8b6e09a54


```{r}

acs_variables_2021 <- load_variables(2021, "acs5", cache = TRUE) |> 
  mutate(label = str_to_lower(label),
         concept = str_to_lower(concept))

glimpse(acs_variables_2021)

```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2021 5-year estimates? You will need to find the variable for median age first. 


**Q1.1** Find variable for median age

**A1.1**

*B01002_001 = variable*
label = estimate!!median age --!!total: 
concept = median age by sex
geog = block


```{r}

#filter label column by median and age

acs_variables_2021_median_age_results <- acs_variables_2021 |> 
  filter(
    str_detect(label, "median"),
    str_detect(label, "age"),
  )

```


```{r}

#filter concept column to remove "alone" 
acs_variables_2021_median_age_results |> 
  filter(
    !str_detect(concept, "alone")
  )

#B01002_001 is the variable (I think)

```

**Q1.2** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2021 5-year estimates? 

**A1.2**

Baltimore city

```{r}
#AC Question: why are we calling the county in geography/why doesn't this work when we call "block" instead? Esp since "block"  is the geog type given to the variable "B01002_001"

md <- get_acs(geography ="county",
              variables = c(median_age = "B01002_001"),
              state = "MD",
              year = 2021) |> 
  arrange(estimate + moe)

md

```
```


**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. 

**A2** 

Talbot County and Worcester County


```{r}

md |> 
  clean_names() |>
  filter(estimate >= ("50")) |> 
arrange(estimate)

```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. 

To do this, you'll need to:

1. Load the Montgomery County 911 overdose calls data.

```{r}
montgomery_od_2022 <- read_csv('/Users/AlishaCamacho/Documents/GitHub/camacho-alisha-jour-772/lab_07/data/montgomery_2022_overdoses.csv')
```

2. Create a new dataframe with the total number of calls for each zip code.

```{r}
#clean data first

montgomery_calls <- montgomery_od_2022 |> 
  clean_names() |> 
  select(city, state, zip, incident_id) |> 
  mutate(city = str_to_lower(city)) |> 
  mutate(state = str_to_lower(state)) |> 
  mutate(zip = as.character(zip) |> trimws())
  
  montgomery_calls

```

```{r}

#calls for each zip 

montgomery_calls_zip <- montgomery_calls |> 
  group_by(zip) |>
  mutate(count = n()) |>
  summarise(calls = sum(count)) |> 
  arrange(calls)

montgomery_calls_zip

```


3. Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one).

#answer to part 3: 
 variable = B09021_001
 label = estimate!!total:
 concept = living arrangements of adults 18 years and over by age


```{r}
#filter by label

acs_variables_2021_label <- acs_variables_2021 |> 
  filter(!str_detect(label, "under 5 years|5 to 9 years|10 to 14 years|15 to 17 years|18 to 19 years|20 to 24 years|25 to 29 years|30 to 34 years|18 and 19 years|35 to 44 years|45 to 54 years|55 to 64 years|65 to 74 years|75 to 84 years|85 years and older|male|female|sex|race|american indian|black|white|asian|median age|native|hispanic or latino|americas|asia"))
   
acs_variables_2021_label

```

```{r}
#filter by concept 
#continue search for total 18+ population

acs_variabels_2021_label_concept <- acs_variables_2021_label |> 
  filter(!str_detect(concept, "alone|two or more races|hispanic or latino|sex|puerto rico|6 to 17 years|place of birth|geographical mobility|means of transportation to work|first marriage|a birth|language spoken at home|poverty|disability|income|employment|work|room|mortgage|tenure|female|industry|median age|insurance|occupation|ancestry|under 18|children|citizenship status|citizen|naturalization|nativity|movers|vehicles"))

acs_variabels_2021_label_concept

#B09021_001
#label = estimate!!total:
#concept = living arrangements of adults 18 years and over by age
#geography = block group

```


4. Get from the ACS data from all Maryland zip codes using that variable.

```{r}

md_pop_over_18 <- get_acs(
        geography = "block group",
        variables = "B09021_001",
        state = "MD",
        year = 2021)
md_pop_over_18
```

5. Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.

```{r}

#ACS data frame = md_pop_over_18
#Montgomery County zip code data frame = montgomery_calls_zip
#clean md_pop_over_18 data frame

adult_pop_md <- md_pop_over_18 |> 
  clean_names() |> 
  mutate(zip = str_sub(geoid, 1, 6)) |>
  rename(pop_count = estimate) |> 
  mutate(name = str_to_lower(name)) 

adult_pop_md
#montgomery_calls_zip

```

#reposting question 

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. 
*noting that the Q later specifies we are interested in the adult population only 

#Join dataframes
adult_pop_md
montgomery_calls_zip


#I am at a loss with what I am doing wrong, as I keep getting zero results. Here is one of the many chats i've had with chat gpt on this one: https://chat.openai.com/share/c063b620-4ced-4b18-8f9c-9313e5cedfee

```{r}

adult_pop_md  |>  
  inner_join(montgomery_calls_zip)

```


```{r}
#montgomery_calls_adult_pop <- inner_join(montgomery_calls_zip, adult_pop_md, by = "zip")

#non_matching_rows <- anti_join(montgomery_calls_zip, adult_pop_md, by = "zip")

```


6. Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.

Which zip code has the highest rate of calls? Which has the lowest?

**A3** 

```{r}



```

**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?



**A4** 

20860 is the answer


**Q5** Go back to your joined dataframe from Q3 and add two more columns representing the minimum and maximum populations based on the estimate and margin of error, then calculate per capita rates for each. Do you get the same two zip codes for the highest and lowest as you did before? Does that give you more or less confidence in your answers to Q3?


**A5** 

```{r}



```

rm(list = ls())

