---
title: "lab_09.Rmd"
author: "derek willis"
date: "2023-04-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## You will need

* A Datawrapper account
* Data in CSV files in the `data` folder

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}

library(tidyverse)
library(refinr)
library(janitor)
library(lubridate)

rm(list = ls())

```

## Questions

**Q1.** You want to explore the differences between the electoral performances of Dan Cox in 2022 and Larry Hogan in 2018. Except a map is probably not the best way to show that. Let's make a chart instead, and the type is up to you, but the goal is to clearly show the differences between the two.

You MUST do the following things when answering this question:


#Load the md_gov_county.csv and md_gov_county_18.csv files from the data folder

```{r}
#load and clean files

md_gov_county <- read.csv("/Users/AlishaCamacho/Documents/GitHub/camacho-alisha-jour-772/lab_09/data/md_gov_county.csv")

md_gov_county <- md_gov_county |> 
  clean_names() |> 
  mutate(cox = as.numeric(cox)) |> 
  mutate(moore = as.numeric (moore)) |> 
  mutate(lashar = as.numeric(lashar)) |> 
  mutate(wallace = as.numeric(wallace)) |> 
  mutate(harding = as.numeric(harding)) |> 
  mutate(sefcik = as.numeric(sefcik)) |> 
  mutate(write_ins = as.numeric(write_ins)) |> 
  mutate(geoid = as.character(geoid))

md_gov_county

```

```{r}
#percentcox

md_gov_county_cox<- md_gov_county |> 
  rowwise() |> 
  mutate(total_votes = sum(cox + moore + lashar + wallace + harding + sefcik + write_ins)) |> 
  mutate(cox_percent = (cox/total_votes)*100)

md_gov_county_cox

```


```{r}
md_gov_county_18 <- read.csv("/Users/AlishaCamacho/Documents/GitHub/camacho-alisha-jour-772/lab_09/data/md_gov_county_18.csv")

md_gov_county_18 <- md_gov_county_18 |> 
  clean_names() |> 
  mutate(geoid = as.character(geoid)) |> 
  mutate(hogan = as.numeric(hogan)) |> 
  mutate(quinn = as.numeric(quinn)) |> 
  mutate(schlakman = as.numeric(schlakman)) |> 
  mutate(write_ins = as.numeric(write_ins)) |> 
  mutate(jealous = as.numeric(jealous))

md_gov_county_18

```

```{r}

#percenthogan

md_hogan_18 <- md_gov_county_18 |> 
  rowwise() |> 
  mutate(total_votes = sum(hogan + jealous + quinn + schlakman + write_ins)) |> 
  mutate(hogan_percent = (hogan/total_votes)*100)

md_hogan_18

```


#Join those dataframes, specifying the join column. 
md_hogan_18
md_gov_county_cox

#Add columns to generate percentages for Cox and Hogan. *already done

```{r}

cox_v_hogan <- left_join(md_hogan_18, md_gov_county_cox, by = "geoid") |> 
 select(geoid, county.x, hogan_percent, cox_percent)  |> 
 mutate(difference = (hogan_percent - cox_percent))

cox_v_hogan

```

```{r}

#Export that joined dataframe to a CSV file using `write_csv`.

write.csv(cox_v_hogan, file = "cox_v_hogan.csv")

```


* Using that exported CSV, make a chart with Datawrapper. What kind of chart is up to you, so experiment with a few kinds of charts to see what answers the question above in the clearest way. Your chart MUST:

* Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.

* Easily readable labels and values

**Publish your chart** https://www.datawrapper.de/_/pO1iX/

**Describe your thought process below - why you chose to make the chart you did - and add the URL of your chart.**

The bullet bar chart made the most sense to me, as it was the chart that allowed me to quickly digest the information by directly comparing the number of votes for Cox compared with Hogan by county and percentage. You can clearly see that Hogan received  more votes than Cox is all counties. That said, I do wish I could see the percent numbers for both bars while hovering over them, which is not an option for this bar chart. 

I also only included information on Cox and Hogan, since the question was interested in comparing these two candidates specifically. The other candidates would be noise, in my opinion, in this context. 

https://www.datawrapper.de/_/pO1iX/


**Q2.** You're interested in seeing how the methods of voting have changed in Maryland since 2018. Using the `md_votes_by_type.csv` that has data from the top of the ticket (governor in 2018 and 2022, president in 2020)

#load data and review

```{r}
 
md_votes_by_type <- read_csv("/Users/AlishaCamacho/Documents/GitHub/camacho-alisha-jour-772/lab_09/data/md_votes_by_type.csv")

md_votes_by_type

```


#Make a chart showing the percentage of votes cast in each category over three years:

- early voting
- election day
- mail 

**AC question** 
#why do 2020 votes account for a total of 96%? Where is the other 4%? 

```{r}
md_votes_percentage <- md_votes_by_type |>
  mutate(Year = as.character(Year)) |> 
  clean_names() |> 
  mutate(percent_early = (early/total)*100) |> 
  mutate(percent_election_day = (election_day/total)*100) |> 
  mutate(percent_mail = (mail/total)*100) |> 
  rowwise(percent_early, percent_election_day, percent_mail) |> 
  mutate(sum = percent_early + percent_election_day + percent_mail) # |> 
# select(year, percent_early, percent_election_day, percent_mail)

md_votes_percentage

```


You'll need to make those percentages in R and then export the CSV using `write_csv`

```{r}

#I give up on this puzzle (changing the headers back to Title_Case) for now

#md_votes_percentage
#colnames(md_votes_percentage) <- c("Year, early_x, election_day_x, mail_x, total_x, Early, Election Day, Mail, sum")

#md_votes_percentage


write_csv(md_votes_percentage, file = "md_votes_percentage.csv")

```

Then make the chart in Datawrapper. Have a: 
- title
- description
- source that tells readers about the data. 

The title and description should highlight the most newsworthy or important aspect.
HINT: In Datawrapper's "Check and Describe" tab, you can click on a column to get the option of "hiding" it in any visualization.

What do you think the story here is? Put the answer below along with the URL of your chart.


**A2.** 
https://www.datawrapper.de/_/arzux/

I think the story is learning why people chose not to do  mail-in ballots again in 2022, after surging during the pandemic. Of course, the reason it surged in the first place is related to the pandemic, but would want to know why people elected to vote in person on election day in 2022. 

##

**Q3.** You want to know more about the geography of 911 overdose calls in Montgomery County, Maryland, in 2022. To do that, you'll make a choropleth map in Datawrapper. You will need to:

* Read the file into R and create columns for:
- the number of calls in each zip code 
- the percentage of all calls in each zip code 

*then export that CSV to use in your map.

```{r}

mc_od <- read_csv("/Users/AlishaCamacho/Documents/GitHub/camacho-alisha-jour-772/lab_09/data/montgomery_2022_overdoses.csv") |> 
  select(zip, incident_id) |> 
  mutate(zip = as.character(zip)) |> 
  group_by(zip) |> 
  summarise(count = n()) |> 
  mutate(percent = (count/sum(count))*100) |> 
  mutate(percent_rounded = round(percent)) |> 
  mutate(percent_total = sum(percent_rounded)) |> 
  select(zip, count, percent_rounded) 

mc_od 

write_csv(mc_od, file = "mc_od.csv")

```


* Search for Maryland zip codes (use the 2020 data) as the base map.
* Upload the CSV file.
* Change the tooltip to show both the number and the percentage (you can put them side by side or put one under the other). Make sure to properly format both numbers to make them readable.

* Add labels to your map that locate prominent places in the county.

* Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.

* Make any other changes to data that helps you make a more accurate map.

What do you think the story here is, and what would you want to know more about? Put the answer below along with the URL of your map.

**A3.** 
https://www.datawrapper.de/_/tvl2g/


I am not sure what the story is here, based on this map alone. I'd need more information to overlay, including population size per zip code, medium income, residential vs. business districts, rehab clinics, crime, park space, etc. The best I got is that the most calls occurred in the zipcode with Germantown, but then that brings into question whether Germantown specifically had high levels of calls, and why or why not. I'd need to see the data broken down by census tract or otherwise to find a meaningful story.  



